Regression Models
=================
*This note is a reorganization of Dr. Brian Caffo's lecture notes for the Coursera course [Regression Models](https://class.coursera.org/regmods-002).*

# Module III : Generalized Linear Models

## Introduction

The Generalized Linear Model (GLM) was introduced in a 1972 RSSB paper by Nelder and Wedderburn. It involves three components

* An *exponential family* model for the response.
* A systematic component via a linear predictor.
* A link function that connects the means of the response to the linear predictor.

> GLMs can be thought of as a two-stage modeling approach. We first model the response variable using a probability distribution, such as the binomial or Poisson distribution. Second, we model the parameter of the distribution using a collection of prediction and a special form of multi regression. - [OpenIntro](http://www.openintro.org)

Mathematically, first we assume that $Y_i \sim N(\mu_i, \sigma^2)$ (the Gaussian distribution is an exponential family distribution) and define the linear predictor to be $\eta_i = \sum_{k=1}^p X_{ik} \beta_k$. The link function is $g(\mu) = \eta$. 

For linear models $g(\mu) = \mu$ so that $\mu_i = \eta_i$. This yields the same likelihood model as our additive error Gaussian linear model $Y_i = \sum_{k=1}^p X_{ik} \beta_k + \epsilon_{i}$ where $\epsilon_i \stackrel{iid}{\sim} N(0, \sigma^2)$.

### Logistic Regression
logistic regression is a type of GLM where regular multiple regression does not work well. Assume that $Y_i \sim Bernoulli(\mu_i)$ so that $E[Y_i] = \mu_i$ where $0\leq \mu_i \leq 1$.
* Linear predictor $\eta_i = \sum_{k=1}^p X_{ik} \beta_k$
* Link function 
$g(\mu) = \eta = \log\left( \frac{\mu}{1 - \mu}\right)$
$g$ is the (natural) log odds, referred to as the **logit**. For a probability p, the function $p/(1 − p)$ is called the **odds ratio**.
* Note then we can invert the logit function as
$$
\mu_i = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)} ~~~\mbox{and}~~~
1 - \mu_i = \frac{1}{1 + \exp(\eta_i)}
$$
Thus the likelihood is
$$
\prod_{i=1}^n \mu_i^{y_i} (1 - \mu_i)^{1-y_i}
= \exp\left(\sum_{i=1}^n y_i \eta_i \right)
\prod_{i=1}^n (1 + \eta_i)^{-1}
$$

### Poisson Regression
Assume that $Y_i \sim Poisson(\mu_i)$ so that $E[Y_i] = \mu_i$ where $0\leq \mu_i$
* Linear predictor $\eta_i = \sum_{k=1}^p X_{ik} \beta_k$
* Link function 
$g(\mu) = \eta = \log(\mu)$
* Recall that $e^x$ is the inverse of $\log(x)$ so that 
$$
\mu_i = e^{\eta_i}
$$
Thus, the likelihood is
$$
\prod_{i=1}^n (y_i !)^{-1} \mu_i^{y_i}e^{-\mu_i}
\propto \exp\left(\sum_{i=1}^n y_i \eta_i - \sum_{i=1}^n \mu_i\right)
$$

### Summary
In each case, the only way in which the likelihood depends on the data is through 
$$\sum_{i=1}^n y_i \eta_i =
\sum_{i=1}^n y_i\sum_{k=1}^p X_{ik} \beta_k = 
\sum_{k=1}^p \beta_k\sum_{i=1}^n X_{ik} y_i
$$
Thus if we don't need the full data, only $\sum_{i=1}^n X_{ik} y_i$. This simplification is a consequence of chosing so-called **canonical** link functions.

All models acheive their maximum at the root of the so called normal equations
$$
0=\sum_{i=1}^n \frac{(Y_i - \mu_i)}{Var(Y_i)}W_i
$$
where $W_i$ are the derivative of the inverse of the link function.

The variances in the GLMs are

* For the linear model $Var(Y_i) = \sigma^2$ is constant.
* For Bernoulli case $Var(Y_i) = \mu_i (1 - \mu_i)$
* For the Poisson case $Var(Y_i) = \mu_i$. 
* In the latter cases, it is often relevant to have a more flexible variance model, even if it doesn't correspond to an actual likelihood
$$
0=\sum_{i=1}^n \frac{(Y_i - \mu_i)}{\phi \mu_i (1 - \mu_i ) } W_i ~~~\mbox{and}~~~
0=\sum_{i=1}^n \frac{(Y_i - \mu_i)}{\phi \mu_i} W_i
$$
* These are called **quasi-likelihood** normal equations 

### Odds and ends
* The normal equations have to be solved iteratively. Resulting in $\hat \beta_k$ and, if included, $\hat \phi$.
* Predicted linear predictor responses can be obtained as $\hat \eta = \sum_{k=1}^p X_k \hat \beta_k$
* Predicted mean responses as $\hat \mu = g^{-1}(\hat \eta)$
* Coefficients are interpretted as 
$$
g(E[Y | X_k = x_k + 1, X_{\sim k} = x_{\sim k}]) - g(E[Y | X_k = x_k, X_{\sim k}=x_{\sim k}]) = \beta_k
$$
or the change in the link function of the expected response per unit change in $X_k$ holding other regressors constant.
* Variations on Newon/Raphson's algorithm are used to do it.
* Asymptotics are used for inference usually. 
* Many of the ideas from linear models can be brought over to GLMs.

---
## Logistic Regression
In this section, we will explore the Baltimore Ravens data. First let's load it and check its structure
```{r loadRavens,cache=TRUE}
download.file("https://dl.dropboxusercontent.com/u/7710864/data/ravensData.rda"
              , destfile="../03_02_binaryOutcomes/data/ravensData.rda", method = "wget")
load("../03_02_binaryOutcomes/data/ravensData.rda")
head(ravensData)
str(ravensData)
```

### Linear Regression
The linear regression can be modeled as
$$ RW_i = b_0 + b_1 RS_i + e_i $$
Where
* $RW_i$ - 1 if a Ravens win, 0 if not
* $RS_i$ - Number of points Ravens scored
* $b_0$ - probability of a Ravens win if they score 0 points
* $b_1$ - increase in probability of a Ravens win for each additional point
* $e_i$ - residual variation due 


```{r linearReg, dependson = "loadRavens", cache=TRUE}
lmRavens <- lm(ravensData$ravenWinNum ~ ravensData$ravenScore)
summary(lmRavens)$coef
```
### Odds

__Binary Outcome 0/1__

$$RW_i$$  

__Probability (0,1)__

$$\rm{Pr}(RW_i | RS_i, b_0, b_1 )$$


__Odds $(0,\infty)$__
$$\frac{\rm{Pr}(RW_i | RS_i, b_0, b_1 )}{1-\rm{Pr}(RW_i | RS_i, b_0, b_1)}$$ 

__Log odds $(-\infty,\infty)$__

$$\log\left(\frac{\rm{Pr}(RW_i | RS_i, b_0, b_1 )}{1-\rm{Pr}(RW_i | RS_i, b_0, b_1)}\right)$$ 


### Linear vs. logistic regression

__Linear__

$$ RW_i = b_0 + b_1 RS_i + e_i $$

or

$$ E[RW_i | RS_i, b_0, b_1] = b_0 + b_1 RS_i$$

__Logistic__

$$ \rm{Pr}(RW_i | RS_i, b_0, b_1) = \frac{\exp(b_0 + b_1 RS_i)}{1 + \exp(b_0 + b_1 RS_i)}$$

or

$$ \log\left(\frac{\rm{Pr}(RW_i | RS_i, b_0, b_1 )}{1-\rm{Pr}(RW_i | RS_i, b_0, b_1)}\right) = b_0 + b_1 RS_i $$

### Logistic Regression

$$ \log\left(\frac{\rm{Pr}(RW_i | RS_i, b_0, b_1 )}{1-\rm{Pr}(RW_i | RS_i, b_0, b_1)}\right) = b_0 + b_1 RS_i $$


* $b_0$ - Log odds of a Ravens win if they score zero points
* $b_1$ - Log odds ratio of win probability for each point scored (compared to zero points)
* $\exp(b_1)$ - **Odds ratio** of win probability for each point scored (compared to zero points)

### Odds
- Imagine that you are playing a game where you flip a coin with success probability $p$.
- If it comes up heads, you win $X$. If it comes up tails, you lose $Y$.
- What should we set $X$ and $Y$ for the game to be fair?

    $$E[earnings]= X p - Y (1 - p) = 0$$
- Implies
    $$\frac{Y}{X} = \frac{p}{1 - p}$$    
- The odds can be said as "How much should you be willing to pay for a $p$ probability of winning a dollar?"
    - (If $p > 0.5$ you have to pay more if you lose than you get if you win.)
    - (If $p < 0.5$ you have to pay less if you lose than you get if you win.)

```{r logReg, dependson = "loadRavens", fig.width = 5, fig.height = 5}
logRegRavens <- glm(ravensData$ravenWinNum ~ ravensData$ravenScore,family="binomial")
summary(logRegRavens)
plot(ravensData$ravenScore,logRegRavens$fitted,pch=19,col="blue",xlab="Score",ylab="Prob Ravens Win")
```
**Odds ratios and confidence intervals**

```{r dependson = "logReg",fig.height=4,fig.width=4}
exp(logRegRavens$coeff)
exp(confint(logRegRavens))
```

**ANOVA for logistic regression**

```{r dependson = "logReg",fig.height=4,fig.width=4}
anova(logRegRavens,test="Chisq")
```

We can interpreting odds ratios

* Not probabilities 
* Odds ratio of 1 = no difference in odds
* Log odds ratio of 0 = no difference in odds
* Odds ratio < 0.5 or > 2 commonly a "moderate effect"
* Relative risk $\frac{\rm{Pr}(RW_i | RS_i = 10)}{\rm{Pr}(RW_i | RS_i = 0)}$ often easier to interpret, harder to estimate
* For small probabilities RR $\approx$ OR but __they are not the same__!

---
## Poisson Regression
There are many data taking the form of **counts**, such as the number of calls to a call center, the number of flu cases in an area, the number of cars that cross a bridge and etc. The data may also be in the form of **rates**, such as the percentage of children passing a test or the percentage of hits to a website from a country.

In all the above cases, Linear regression with transformation is an option.

First, we recall that the Poisson distribution is a useful model for counts and rates (rate is count per some monitoring time). Some examples uses of the Poisson distribution

* Modeling web traffic hits
* Incidence rates
* Approximating binomial probabilities with small $p$ and large $n$
* Analyzing contigency table data

For the Poisson distribution
- $X \sim Poisson(t\lambda)$ if
$$
P(X = x) = \frac{(t\lambda)^x e^{-t\lambda}}{x!}
$$
For $x = 0, 1, \ldots$.
- The mean of the Poisson is $E[X] = t\lambda$, thus $E[X / t] = \lambda$
- The variance of the Poisson is $Var(X) = t\lambda$.
- The Poisson tends to a normal as $t\lambda$ gets large.

```{r simPois,fig.height=3,fig.width=10, cache=TRUE}
par(mfrow = c(1, 3))
plot(0 : 10, dpois(0 : 10, lambda = 2), type = "h", frame = FALSE)
plot(0 : 20, dpois(0 : 20, lambda = 10), type = "h", frame = FALSE)
plot(0 : 200, dpois(0 : 200, lambda = 100), type = "h", frame = FALSE) 
```
We can show the the mean and variance of a poisson distribution are equal
```{r}
x <- 0 : 10000; lambda = 3
mu <- sum(x * dpois(x, lambda = lambda))
sigmasq <- sum((x - mu)^2 * dpois(x, lambda = lambda))
c(mu, sigmasq)
```

### Example: Leek Group Website Traffic
As an example, let's consider the daily counts to [Jeff Leek's website](http://biostat.jhsph.edu/~jleek/), since the unit of time is always one day, set $t = 1$ and then the Poisson mean is interpretted as web hits per day. (If we set $t = 24$, it would be web hits per hour). ([http://skardhamar.github.com/rga/](http://skardhamar.github.com/rga/))

```{r leekLoad,cache=TRUE}
download.file("https://dl.dropboxusercontent.com/u/7710864/data/gaData.rda",
              destfile="../03_03_countOutcomes/data/gaData.rda", method = "wget")
load("../03_03_countOutcomes/data/gaData.rda")
gaData$julian <- julian(gaData$date)
str(gaData)
```
If we fit the data with a simple linear regression model

$$ NH_i = b_0 + b_1 JD_i + e_i $$

* $NH_i$ - number of hits to the website
* $JD_i$ - day of the year (Julian day)
* $b_0$ - number of hits on Julian day 0 (1970-01-01)
* $b_1$ - increase in number of hits per unit day
* $e_i$ - variation due to everything we didn't measure


We can also plot the Linear regression line

```{r linReg, dependson="leekLoad",fig.height=4,fig.width=8, cache=TRUE}
plot(gaData$julian,gaData$visits,pch=19,col="darkgrey",xlab="Julian",ylab="Visits")
lm1 <- lm(gaData$visits ~ gaData$julian)
abline(lm1,col="red",lwd=3)
```

Aside, Let's take the natural log of the outcome has a specific interpretation. Consider the model

$$ \log(NH_i) = b_0 + b_1 JD_i + e_i $$

* $NH_i$ - number of hits to the website
* $JD_i$ - day of the year (Julian day)
* $b_0$ - log number of hits on Julian day 0 (1970-01-01)
* $b_1$ - increase in log number of hits per unit day
* $e_i$ - variation due to everything we didn't measure

To exponentiate coefficients
- $e^{E[\log(Y)]}$ geometric mean of $Y$. 
    - With no covariates, this is estimated by $e^{\frac{1}{n}\sum_{i=1}^n \log(y_i)} = (\prod_{i=1}^n y_i)^{1/n}$
- When you take the natural log of outcomes and fit a regression model, your exponentiated coefficients
estimate things about geometric means.
- $e^{\beta_0}$ estimated geometric mean hits on day 0
- $e^{\beta_1}$ estimated relative increase or decrease in geometric mean hits per day
- There's a problem with logs with you have zero counts, adding a constant works
```{r}
round(exp(coef(lm(I(log(gaData$visits + 1)) ~ gaData$julian))), 5)
```

### Linear vs. Poisson regression

__Linear__

$$ NH_i = b_0 + b_1 JD_i + e_i $$

or

$$ E[NH_i | JD_i, b_0, b_1] = b_0 + b_1 JD_i$$

__Poisson/log-linear__

$$ \log\left(E[NH_i | JD_i, b_0, b_1]\right) = b_0 + b_1 JD_i $$

or

$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 + b_1 JD_i\right) $$

**Multiplicative differences**
$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 + b_1 JD_i\right) $$
$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 \right)\exp\left(b_1 JD_i\right) $$

If $JD_i$ is increased by one unit, $E[NH_i | JD_i, b_0, b_1]$ is multiplied by $\exp\left(b_1\right)$

```{r poisReg, dependson="linReg",fig.height=5,fig.width=10, cache=TRUE}
par(mfrow = c(1, 2))
plot(gaData$julian,gaData$visits,pch=19,col="darkgrey",xlab="Julian",ylab="Visits")
glm1 <- glm(gaData$visits ~ gaData$julian,family="poisson")
abline(lm1,col="red",lwd=3); lines(gaData$julian,glm1$fitted,col="blue",lwd=3)
plot(glm1$fitted,glm1$residuals,pch=19,col="grey",ylab="Residuals",xlab="Fitted", main ="Mean-variance relationship")
```

### Model agnostic standard errors 

```{r agnostic}
library(sandwich)
confint.agnostic <- function (object, parm, level = 0.95, ...)
{
    cf <- coef(object); pnames <- names(cf)
    if (missing(parm))
        parm <- pnames
    else if (is.numeric(parm))
        parm <- pnames[parm]
    a <- (1 - level)/2; a <- c(a, 1 - a)
    pct <- stats:::format.perc(a, 3)
    fac <- qnorm(a)
    ci <- array(NA, dim = c(length(parm), 2L), dimnames = list(parm,
                                                               pct))
    ses <- sqrt(diag(sandwich::vcovHC(object)))[parm]
    ci[] <- cf[parm] + ses %o% fac
    ci
}
```
[http://stackoverflow.com/questions/3817182/vcovhc-and-confidence-interval](http://stackoverflow.com/questions/3817182/vcovhc-and-confidence-interval)

### Estimating confidence intervals

```{r}
confint(glm1)
confint.agnostic(glm1)
```

### Rates 

$$ E[NHSS_i | JD_i, b_0, b_1]/NH_i = \exp\left(b_0 + b_1 JD_i\right) $$
$$ \log\left(E[NHSS_i | JD_i, b_0, b_1]\right) - \log(NH_i)  =  b_0 + b_1 JD_i $$
$$ \log\left(E[NHSS_i | JD_i, b_0, b_1]\right) = \log(NH_i) + b_0 + b_1 JD_i $$


```{r ratesFit,dependson="agnostic", cache=TRUE,fig.height=5,fig.width=10}
par(mfrow = c(1, 2))
glm2 <- glm(gaData$simplystats ~ julian(gaData$date),offset=log(visits+1),
            family="poisson",data=gaData)
plot(julian(gaData$date),glm2$fitted,col="blue",pch=19,xlab="Date",ylab="Fitted Counts")
points(julian(gaData$date),glm1$fitted,col="red",pch=19)

glm2 <- glm(gaData$simplystats ~ julian(gaData$date),offset=log(visits+1),
            family="poisson",data=gaData)
plot(julian(gaData$date),gaData$simplystats/(gaData$visits+1),col="grey",xlab="Date",
     ylab="Fitted Rates",pch=19)
lines(julian(gaData$date),glm2$fitted/(gaData$visits+1),col="blue",lwd=3)
```


 










---
## Further resources
* [Wikipedia on Odds Ratio](http://en.wikipedia.org/wiki/Odds_ratio)
* [Wikipedia on Logistic Regression](http://en.wikipedia.org/wiki/Logistic_regression)
* [Logistic regression and glms in R](http://data.princeton.edu/R/glms.html)
* Brian Caffo's lecture notes on: [Simpson's paradox](http://ocw.jhsph.edu/courses/MethodsInBiostatisticsII/PDFs/lecture23.pdf), [Case-control studies](http://ocw.jhsph.edu/courses/MethodsInBiostatisticsII/PDFs/lecture24.pdf)
* [Open Intro Chapter on Logistic Regression](http://www.openintro.org/stat/down/oiStat2_08.pdf)
* [Log-linear models and multiway tables](http://ww2.coastal.edu/kingw/statistics/R-tutorials/loglin.html)
* [Wikipedia on Poisson regression](http://en.wikipedia.org/wiki/Poisson_regression), [Wikipedia on overdispersion](http://en.wikipedia.org/wiki/Overdispersion)
* [Regression models for count data in R](http://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf)
* [pscl package](http://cran.r-project.org/web/packages/pscl/index.html) - the function _zeroinfl_ fits zero inflated models.

---
Previous Module. [Module II : Multivariable Regression](http://rpubs.com/sialy/regmod-mod-2)
